---
title: "Simulation Study: Comparing LATE Estimators"
author: "princeBART authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study: Comparing LATE Estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 5
)
.extdata <- function(fname) {
  f <- system.file("extdata", fname, package = "princeBART")
  if (nzchar(f) && file.exists(f)) return(f)
  vdir <- tryCatch(
    dirname(normalizePath(knitr::current_input())),
    error = function(e) getwd()
  )
  f <- normalizePath(file.path(vdir, "..", "inst", "extdata", fname), mustWork = FALSE)
  if (file.exists(f)) return(f)
  stop("Cannot find extdata/", fname)
}
```

## Overview

This vignette illustrates the finite-sample performance of **princeBART**
in comparison with commonly used estimators for complier-specific causal effects
(often termed LATE) under noncompliance or treatment uptake endogeneity.

The goal of the simulation is not to provide an exhaustive evaluation, but
to highlight how different estimators behave across data-generating processes
that vary in complexity, overlap, and treatment effect heterogeneity. In
particular, the simulation contrasts settings in which linear methods such as
two-stage least squares (2SLS) are well specified with settings in which
nonlinearities and heterogeneous effects are present.

The results shown here are based on a reduced number of simulation iterations
to keep vignette build times reasonable. The full simulation study, including
additional designs and substantially more repetitions, is reported in the
accompanying paper.

The following estimators are compared:

- **2SLS**: Two-stage least squares with robust standard errors  
- **κ-weighting**: Normalized kappa-weighting estimator  
- **PStrata**: Bayesian principal stratification  
- **Prince BART**: Principal stratification with BART (this package)

## Simulation design

The data-generating processes considered here follow the simulation designs
used in the princeBART paper and related methodological work. They are not
constructed to favor princeBART, but instead represent settings previously used
to evaluate instrumental variable estimators under noncompliance, varying
in the degree of overlap, functional form, and treatment effect heterogeneity.

The designs differ in how treatment uptake depends on covariates and the
instrument, and in whether the complier treatment effect is constant or
nonlinear in covariates.

## Data Generating Process

| Design | Compliance (γ) | Treatment Effect μ₁(x) | Propensity f(x) |
|--------|----------------|------------------------|-----------------|
| A | (0, 0, 4) | Constant | Linear: 2x − 1 |
| C | (−1, 2, 2.122) | Nonlinear: (3(x+3))² | Linear: 2x − 1 |

The **δ** parameter controls overlap: smaller δ yields more extreme propensity scores.

```{r load-packages}
library(princeBART)
library(MASS)
library(ivreg)
library(sandwich)
library(kappalate)
library(PStrata)
library(rstan)
library(matrixStats)
library(abind)
library(ggplot2)
library(tidyverse)
library(future)
library(future.apply)
```

### DGP setup

```{r dgp-setup}
# Correlation structure for error terms
rho   <- 0.5
SIGMA <- matrix(c(1, 0, 0, 0, 1, rho, 0, rho, 1), 3, 3)
mu    <- c(0, 0, 0)

# Data generating function
create_data <- function(design, del, n) {
  if (design == "A") {
    gamma <- c(0, 0, 4)
    f_mu1 <- \(x) dnorm(0, 0, 1)
    f_zx  <- \(x) (2 * x - 1)  # linear propensity
  } else if (design == "B") {
    gamma <- c(-1, 2, 2.122)
    f_mu1 <- \(x) dnorm(0, 0, 1)
    f_zx  <- \(x) (2 * x - 1)  # linear propensity
  } else if (design == "C") {
    gamma <- c(-1, 2, 2.122)
    f_mu1 <- \(x) (3 * (x + 3))^2  # nonlinear effect
    f_zx  <- \(x) (2 * x - 1)      # linear propensity
  } else if (design == "D") {
    gamma <- c(-1, 2, 2.122)
    f_mu1 <- \(x) (3 * (x + 3))^2  # nonlinear effect
    f_zx  <- \(x) (x + x^2 - 1)    # nonlinear propensity
  } else {
    stop("Unknown design")
  }

  g0   <- log((1 - del) / del)
  f_pz <- \(x) 1 / (1 + exp(-f_zx(x) * g0))
  f_gd <- \(x, z) gamma[1] + x * gamma[2] + z * gamma[3]

  x  <- runif(n)
  pz <- f_pz(x)
  z  <- (runif(n) < pz) |> as.integer()
  E  <- MASS::mvrnorm(n, mu, SIGMA)

  # Potential treatments (principal strata)
  d1 <- f_gd(x, 1) > E[, 3]
  d0 <- f_gd(x, 0) > E[, 3]
  d  <- d1 * z + (1 - z) * d0

  # Potential outcomes
  y0      <- 0 > E[, 1]
  y1_star <- f_mu1(x) + E[, 2]
  y1      <- y1_star > median(y1_star)
  y       <- d * y1 + (1 - d) * y0

  # True LATE
  late <- mean((y1 - y0) * (d1 - d0)) / mean(d1 - d0)
  data <- data.frame(Y = y, X = x, W = d, Z = z)
  list(data = data, late = late)
}
```

## Estimator Implementations

### 2SLS

```{r ivfit-fitter}
ivfit_fitter <- function(data) {
  ivfit <- ivreg(Y ~ X + W, ~ X + Z, data = data)
  a <- coef(summary(ivfit, vcov = sandwich::sandwich, df = Inf))["W", 1:2]
  b <- confint(ivfit, vcov = sandwich::sandwich, df = Inf)["W", ]
  c(a, b)
}
```

### κ-weighting (normalized)

```{r cbps-fitter}
cbps_fitter <- function(data) {
  cbps <- kappalate(Y ~ X | W | Z, data)
  est  <- cbps$coefficients |> as.vector()
  se   <- cbps$vcov_matrix |> sqrt() |> as.vector()
  ci   <- c(-1.96, 1.96) * se + est
  c(est, se, ci)
}
```


### PStrata

```{r PStrata-fitters}
# PStrata model object
ps_obj <- function(data) {
  s_formula <- formula(Z + W ~ X)
  y_formula <- update(s_formula, Y ~ .)
  PSObject(
    S.formula = s_formula,
    Y.formula = y_formula,
    prior_intercept = prior_normal(),
    Y.family = binomial(),
    data = data,
    strata = c(n = "00*", c = "01", a = "11*")
  )
}

# Pre-compile Stan model (run once)
pstrata_compile <- function(data) {
  psobj <- ps_obj(data)
  ps_fit <- PStrata(psobj, cores = 1, chains = 1, iter = 20)
  ps_fit$post_samples
}

# Extract fitted values from PStrata fit
fitted_pstrata <- function(ps_fit) {
  beta_s <- rstan::extract(ps_fit$post_samples)$beta_S
  beta_g <- rstan::extract(ps_fit$post_samples)$beta_G
  psdt <- PStrata::make_standata(ps_fit$PSobject)

  # Fitted outcome probabilities
  mu <- apply(beta_g, 1:2, \(b) plogis(psdt$XG %*% b))

  # Fitted stratum probabilities
  pi <- lapply(seq_len(nrow(beta_s)), \(i) {
    psdt$XS %*% t(rbind(0, beta_s[i, , ])) |>
      apply(1, \(v) v - matrixStats::logSumExp(v)) |> exp() |> t()
  }) |> abind::abind(along = 3) |> aperm(c(1, 3, 2))

  dimnames(mu)[[3]] <- c("n", "c0", "c1", "a")
  dimnames(pi)[[3]] <- c("n", "c", "a")
  list(pi, mu)
}

# Compute mixed average complier effect
mixed_complier_effect <- function(fitval) {
  m1_c <- fitval[[2]][, , "c1"]
  m0_c <- fitval[[2]][, , "c0"]
  pi_c <- fitval[[1]][, , "c"]
  est <- apply((m1_c - m0_c) * pi_c, 2, sum) / apply(pi_c, 2, sum)
  c(mean(est), sd(est), quantile(est, c(.025, .975)))
}

# PStrata fitter (uses pre-compiled model)
pstrata_fitter <- function(data, mcmc_iters = 200, pre_fit = NULL) {
  psobj <- ps_obj(data)
  ps_fit <- PStrata(
    psobj, cores = 1, chains = 1, iter = mcmc_iters, fit = pre_fit
  )
  mixed_complier_effect(fitted_pstrata(ps_fit))
}


```

### Prince BART

```{r princebart-fitters}

late_draws_from_chains <- function(chains) {
  unlist(lapply(chains, function(chain) {
    d <- chain$m_y1c - chain$m_y0c
    p_c <- 1 - chain$p_a - chain$p_n
    rowSums(d * p_c) / rowSums(p_c)
  }), use.names = FALSE)
}


princeb_fitter <- function(data, mcmc_iters = 200) {
  fit <- princeBART::prince_BART(
    formula = Y ~ X | Z | W,
    data = data,
    n_warmup = mcmc_iters,
    n_samples = mcmc_iters,
    n_chains = 1,
    keep_trees = FALSE
  )
  draws <- late_draws_from_chains(fit$chains)
  c(mean(draws), sd(draws), quantile(draws, c(0.025, 0.975)))
}


```

## Summary Functions

```{r summary-functions}
# Absolute bias
bias <- function(res) {
  truth <- mean(res$truth)
  apply(res$results[, "est", ], 1, \(x) abs(mean(x - truth)))
}

# Mean squared error
mse <- function(res) {
  truth <- mean(res$truth)
  apply(res$results[, "est", ], 1, \(x) mean((x - truth)^2))
}

# Relative MSE (normalized to 2SLS)
rmse <- function(res) {
  m <- mse(res)
  m / m[1]
}

# Coverage of 95% CI
cvg <- function(res) {
  truth <- mean(res$truth)
  apply(res$results[, c("ci_l", "ci_u"), ], 1,
        \(x) mean(x["ci_l", ] <= truth & truth <= x["ci_u", ]))
}
```

## Run Simulation

```{r run-simulation}
# Simulation parameters (reduced for vignette)
m <- 2          # iterations (paper uses 400)
designs <- c("A", "C")
deltas <- c(0.02, 0.05)
ns <- 400
```


```{r run-simulation-main, message=FALSE, warning=FALSE, results='hide', eval=FALSE }
set.seed(0203)

# Save and restore plan so worker teardown happens before knitr exits.
old_plan <- future::plan()
on.exit(future::plan(old_plan), add = TRUE)

# Heavy Stan/BART calls inside each worker can exhaust memory if worker count
# is too high. Keep this conservative for stable vignette rendering.
workers <- min(2L, future::availableCores())
future::plan(future::multisession, workers = workers)

# Compile PStrata Stan model once
cat("Compiling PStrata model...\n")
compiled_ps <- pstrata_compile(create_data("A", 0.05, 500)$data)

# Results storage
sim_results <- list()

# Main simulation loop
for (design in designs) {
  for (del in deltas) {
    cat(sprintf("Running: Design %s, δ = %s\n", design, del))

    results <- future_lapply(1:m, function(iter) {
      sim <- create_data(design, del, ns)

      late <- matrix(NA, 4, 4, dimnames = list(
        c("2sls", "kappa", "pstrata", "princeb"),
        c("est", "se", "ci_l", "ci_u")
      ))

      late["2sls", ]    <- ivfit_fitter(sim$data)
      late["kappa", ]   <- cbps_fitter(sim$data)
      late["pstrata", ] <- pstrata_fitter(sim$data, 200, compiled_ps)
      late["princeb", ] <- princeb_fitter(sim$data, 200)

      list(estimates = late, truth = sim$late)
    }, future.seed = TRUE)

    # Combine results
    sim_results[[paste(design, del)]] <- list(
      results = abind::abind(lapply(results, \(x) x$estimates), along = 3),
      truth = sapply(results, \(x) x$truth)
    )
  }
}
# To regenerate: saveRDS(sim_results, file = "inst/extdata/sim_results.rds")
```

## Results

### Compile Metrics

```{r compile-metrics}
# Load precomputed results
sim_results <- readRDS(.extdata("sim_results.rds"))

results_df <- expand.grid(design = designs, delta = deltas) |>
  rowwise() |>
  mutate(res = list(sim_results[[paste(design, delta)]])) |>
  mutate(
    rmse_vals = list(rmse(res)),
    cvg_vals = list(cvg(res)),
    bias_vals = list(bias(res))
  ) |>
  unnest(c(rmse_vals, cvg_vals, bias_vals)) |>
  mutate(
    estimator = rep(
      c("2SLS", "κ-weighting", "PStrata", "Prince BART"), n() / 4
    ) |> factor(
      levels = c("2SLS", "κ-weighting", "PStrata", "Prince BART")
    ),
    design_label = factor(design,
      levels = c("A", "C"),
      labels = c("Design A (simple)", "Design C (nonlinear)")
    )
  )
```

### Relative MSE Plot

```{r rmse-plot, fig.height=4}
ggplot(results_df, aes(x = factor(delta), y = rmse_vals,
                        color = estimator, group = estimator)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  facet_wrap(~ design_label) +
  scale_y_log10() +
  labs(
    title = "Relative MSE by Estimator and Design",
    subtitle = "Values < 1 indicate improvement over 2SLS",
    x = "Overlap (δ)",
    y = "Relative MSE (log scale)",
    color = "Estimator"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
```

### Coverage Plot

```{r coverage-plot, fig.height=4}
ggplot(results_df, aes(x = factor(delta), y = cvg_vals,
                        color = estimator, group = estimator)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray50") +
  facet_wrap(~ design_label) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "95% CI Coverage by Estimator and Design",
    subtitle = "Dashed line shows nominal 95% level",
    x = "Overlap (δ)",
    y = "Coverage",
    color = "Estimator"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
```

### Summary Table

```{r summary-table}
results_df |>
  select(design, delta, estimator, rmse_vals, cvg_vals, bias_vals) |>
  rename(
    Design = design,
    `δ` = delta,
    Estimator = estimator,
    `Rel. MSE` = rmse_vals,
    Coverage = cvg_vals,
    `|Bias|` = bias_vals
  ) |>
  knitr::kable(digits = 3)
```

## Interpretation

The simulation illustrates that:


1. **PStrata** performs better than **2SLS** in the linear case (Design A)
2. **κ-weighting** performs better than **2SLS** in the nonlinear case (Design C)
3. **Prince BART** outperforms **2SLS** in both scenarios, showing robustness to design complexity.

These results should be interpreted as illustrative rather than definitive,
and are intended to complement the more extensive simulation results reported
in the paper.
